---
title: "Research project 1: final version"
format: 
  html:
    self_contained: true
    toc: true
    embed-resources: true
editor: visual
author: Chita Chaaban
date: Winter 2026
---

## Introduction draft

People engage with television shows and films not only for entertainment, but also as a means of identifying with fictional characters. Viewers often evaluate characters in relation to themselves, drawing comparisons in terms of values, behaviors, personality traits, and broader aspects of identity. This process of identification is central to how fictional narratives are experienced and interpreted, and it varies across individuals and narrative contexts.

One way to capture this process is through perceived similarity, which refers to the extent to which an individual feels that a fictional character resembles them. The Self-Report Character Personality Quiz (SWCPQ) was designed to measure this dimension of identification. After completing a self-report personality questionnaire, respondents are asked to rate how similar they feel to fictional characters drawn from a wide range of cultural universes, including television series, films, animation, and comics. Similarity is reported on a six-point scale, providing a quantitative measure of identification with fictional characters. In addition to similarity ratings, the dataset includes demographic information such as age group, gender, and native language, as well as personality indicators.

This report investigates whether perceived similarity to fictional characters varies according to the type of fictional universe in which those characters appear. Because fictional narratives differ in their level of realism, the opportunities for viewers to relate to characters and compare themselves to them may also differ. Specifically, it assesses whether characters originating from realistic versus fantastical narratives are associated with different levels of perceived similarity.

Realist universes typically portray everyday social roles, familiar constraints, and plausible behaviors and reactions that closely resemble those encountered in ordinary life. In contrast, fantastical universes often incorporate supernatural elements, alternative rules governing the world, and exaggerated or non-human abilities. These structural differences between narrative worlds may influence the extent to which viewers can engage in self-referencing and perceive fictional characters as similar to themselves. Our hypothesis is that characters grounded in realistic social contexts may therefore facilitate stronger perceived similarity than characters operating in fantastical settings.

Due to the size of the data set and the diversity of fictional universes represented, the analysis is restricted to the 50 most frequently rated universes among the 341 observed, in order to ensure adequate sample sizes and analytic tractability. These universes were subsequently classified into realist and fantastical categories based on predefined, nested criteria related to narrative realism (e.g., plausibility of events, adherence to real-world social rules, presence of supernatural elements). The classification process was supported by AI-assisted validation to ensure consistency.

In addition to a main effect of fictional genre, the analysis considers gender and age group as moderating variables. Identification with fictional characters is not solely determined by narrative features, but may also depend on individual characteristics that shape how narratives are interpreted and experienced. Gender and age are both associated with differences in socialization, media exposure, and identity formation, which can influence how individuals relate to fictional characters and evaluate similarity. As a result, the contrast between realist and fantastical universes may not operate uniformly across groups. The effect of genre on perceived similarity may therefore vary depending on gender and age group, motivating the examination of interaction effects between fictional genre and these individual-level attributes.

The question we are trying to answer is therefore: How does fictional genre (realistic versus fantastical) influence perceived similarity to fictional characters, and does this relationship vary across gender and age groups?

**H1:** *Characters from realist fictional universes are associated with higher perceived similarity scores than characters from fantastical fictional universes.*

Research on media and film preferences consistently shows that gender is systematically associated with genre engagement. Empirical studies find that men report stronger preferences for genres closely associated with fantastical and speculative elements, such as science fiction, fantasy, and action, whereas women report stronger preferences for genres centered on everyday social interactions, such as drama and romance (Märtens & Reinecke, 2021; Redondo & Holbrook, 2010). Because realist narratives more often emphasize interpersonal relationships, plausible social contexts, and emotionally grounded character development, these genres may facilitate stronger self–character comparison among women. Consequently, the perceived similarity advantage of realist over fantastical universes is expected to be larger for women than for men.

**H2:** *The difference in perceived similarity between realist and fantastical universes will be stronger for women than for men.*

Age has been shown to structure media preferences and genre engagement in systematic ways. Studies using survey and audience data demonstrate that genre choices vary across age groups, reflecting differences in life experience, motivations, and the types of narratives viewers find meaningful (Mares & Sun, 2010). Additional evidence suggests that as individuals age, media engagement becomes increasingly oriented toward meaningful, emotionally grounded content rather than novelty or spectacle (Mares et al., 2016). Given that realist universes are more likely to offer socially plausible situations and recognizable life contexts, older respondents may find it easier to relate to characters in these settings. As a result, the realism advantage in perceived similarity is expected to increase with age.

**H3:** *The difference in perceived similarity between realist and fantastical universes will be stronger for older respondents than for younger respondents.*

Dataset:

Beyond perceived similarity ratings, the data set provides several individual-level attributes that are relevant for the analysis. Each respondent is characterized by demographic variables, including age group, gender, and native language, as well as personality indicators such as Enneagram type. In addition, similarity ratings are linked to specific fictional characters and the fictional universes to which those characters belong. Because respondents rate multiple characters generated by an algorithm, and because these characters originate from different fictional universes, part of the data cleaning resulted to a nested structure, with multiple character-level observations associated with the same respondent and universe. This structure allows the analysis to examine variation in perceived similarity both across fictional universes and across individuals.

To make this structure suitable for hypothesis testing, several data preparation steps are required. In particular, the nested nature of the data necessitates a reorganization of the observations to ensure that comparisons across genres and individuals are meaningful. The preparation process therefore focuses on cleaning invalid or missing responses, standardizing key variables, and aggregating similarity ratings at an appropriate level of analysis. These steps allow the construction of a dataset in which each observation reflects a respondent’s average identification within a given fictional universe, while preserving relevant demographic and personality information for subsequent analysis.

## Data Cleaning

The SWCPQ dataset links respondent-level demographic characteristics to multiple similarity ratings generated for fictional characters across a wide range of narrative universes. Because each participant evaluates several characters produced by the quiz algorithm, the data contain repeated observations nested within respondents and fictional universes. This structure enables the analysis to examine how perceived similarity varies across narrative contexts while accounting for individual-level attributes such as gender and age group. To support hypothesis testing, the analysis focuses on constructing a respondent–universe unit of observation that captures average perceived similarity within a given fictional universe, while preserving the demographic variables required to assess moderation effects.

Load packages and import raw data:

```{r}
library(tidyverse)
library(readr)
library(tidyr)
library(stringr)

# Import raw SWCPQ survey data
Data_ID <- read_csv("SWCPQ-Identification-Survey-Dataset-July2022.csv", show_col_types = FALSE)

```

Clean respondent-level variables

```{r}
Data_clean <- Data_ID %>%
  mutate(
    # Recode placeholder values (0) as missing where 0 does not represent a valid response
    gender = na_if(gender, 0),
    engNat = na_if(engNat, 0),

    # Convert categorical variables to factors
    gender = factor(gender),
    engNat = factor(engNat),

    # Relabel coded categories for interpretability
    age_group = factor(
      age_group,
      levels = c(1, 2, 3),
      labels = c("Group 1", "Group 2", "Group 3")
    ),
    screen = factor(
      screen,
      levels = c(1, 2),
      labels = c("Type 1", "Type 2")
    ),

    # Create a stable respondent identifier
    resp_id = row_number()
  )
```

Reshape nested similarity ratings into long format

```{r}
ratings_long <- Data_clean %>%
  # Keep only respondents with recorded similarity ratings
  filter(!is.na(survey_ratings)) %>%
  # Split nested character ratings into one row per rating
  separate_rows(survey_ratings, sep = "\\], \\[") %>%
  mutate(
    # Remove brackets introduced by string encoding
    survey_ratings = str_replace_all(survey_ratings, "\\[|\\]", "")
  ) %>%
  # Extract character ID, similarity score, and response time
  separate(
    survey_ratings,
    into = c("character_id", "similarity", "rating_time"),
    sep = ", "
  ) %>%
  mutate(
    character_id = str_replace_all(character_id, '"', ""),
    similarity = as.numeric(similarity),
    rating_time = as.numeric(rating_time)
  )

```

Link character ratings to fictional universes:

```{r}
ratings_long <- ratings_long %>%
  mutate(
    # Universe identifier is the prefix of character_id
    universe = sub("/.*", "", character_id)
  )

# Keep only ratings from universes selected by the respondent
keep <- grepl(
  paste0("(^|,)", ratings_long$universe, "(,|$)"),
  ratings_long$universes_selected
)

ratings_sel <- ratings_long[keep, ]

```

Aggregate similarity at the respondent–universe level

```{r}
ratings_long <- ratings_long %>%
  mutate(
    # Universe identifier is the prefix of character_id
    universe = sub("/.*", "", character_id)
  )

# Keep only ratings from universes selected by the respondent
keep <- grepl(
  paste0("(^|,)", ratings_long$universe, "(,|$)"),
  ratings_long$universes_selected
)

ratings_sel <- ratings_long[keep, ]

```

Construct the analysis dataset

```{r}
users_base <- Data_clean %>%
  select(
    resp_id,
    year,
    country,
    age_group,
    gender,
    engNat,
    screen
  )

Data_RR1 <- user_universe_summary %>%
  left_join(users_base, by = "resp_id") %>%
  rename(universe_selected = universe)

```

Restrict to top 50 universes and classify genre

```{r}
# Identify the 50 most frequently rated universes
top_50_list <- Data_RR1 %>%
  count(universe_selected, sort = TRUE) %>%
  slice_head(n = 50) %>%
  pull(universe_selected)

Data_RR1 <- Data_RR1 %>%
  filter(universe_selected %in% top_50_list)

```

Map Realist vs Fantastical universes

```{r}
realist_universes <- c(
  "S", "SHL", "FR", "B99", "T7S", "MF", "HIMYM", "NG", "PR",
  "TO", "GA", "ONB", "CM", "GILG", "GLEE", "MG", "CAS",
  "LS", "PC", "PP", "T", "RD", "RAT"
)

fantastical_universes <- c(
  "HP", "MCU", "ALA", "GOT", "ST", "HG", "UA", "XM", "LOTR",
  "SW", "SL", "LU", "DN", "SK", "SBSP", "FUTR", "FN",
  "TS", "F", "BATB", "WO", "GP", "BOBB", "AL"
)

genre_map <- tibble(
  universe_selected = c(realist_universes, fantastical_universes),
  universe_genre = c(
    rep("Realist", length(realist_universes)),
    rep("Fantastical", length(fantastical_universes))
  )
)

Data_RR1 <- Data_RR1 %>%
  left_join(genre_map, by = "universe_selected") %>%
  # Keep only universes with an explicit genre classification
  filter(universe_genre %in% c("Realist", "Fantastical")) %>%
  mutate(universe_genre = factor(universe_genre))

```

## **Descriptive statistics:**

```{r}
view(Data_RR1)
str(Data_RR1)
library(dplyr)
library(ggplot2)

Data_RR1 %>%
  summarise(
    n_obs = n(),
    n_resp = n_distinct(resp_id),
    mean_similarity = mean(similarity, na.rm = TRUE),
    sd_similarity = sd(similarity, na.rm = TRUE),
    mean_n_ratings = mean(n_ratings, na.rm = TRUE)
  )

table(Data_RR1$gender, useNA = "ifany")
table(Data_RR1$age_group, useNA = "ifany")
table(Data_RR1$engNat, useNA = "ifany")
table(Data_RR1$screen, useNA = "ifany")
table(Data_RR1$universe_genre, useNA = "ifany")

Data_RR1 %>%
  group_by(universe_genre) %>%
  summarise(
    mean_similarity = mean(similarity, na.rm = TRUE),
    sd_similarity = sd(similarity, na.rm = TRUE),
    n = n()
  )

Data_RR1 %>%
  group_by(gender) %>%
  summarise(
    mean_similarity = mean(similarity, na.rm = TRUE),
    n = n()
  )

Data_RR1 %>%
  group_by(age_group) %>%
  summarise(
    mean_similarity = mean(similarity, na.rm = TRUE),
    n = n()
  )

ggplot(Data_RR1, aes(x = universe_genre, y = similarity)) +
  geom_boxplot() +
  theme_minimal()

ggplot(Data_RR1, aes(x = age_group, y = similarity)) +
  geom_boxplot() +
  theme_minimal()

ggplot(Data_RR1, aes(x = gender, y = similarity)) +
  geom_boxplot() +
  theme_minimal()

```

Interpretation:

The dataset contains 18,969,784 character ratings from 297,877 respondents, with an overall mean similarity score of 3.95 (SD = 1.50) on a 1–6 scale. On average, respondents rated 4.7 characters per universe, indicating substantial engagement with the task.

Across universe genres, perceived similarity is very close but not identical. Characters from Realist universes show the highest mean similarity (M = 3.97), followed by Fantastical universes (M = 3.95), while Other/Unmapped universes have slightly lower similarity (M = 3.93). The boxplot confirms this pattern: distributions largely overlap, but the median similarity is marginally higher for Realist content, consistent with the hypothesis that realistic settings facilitate identification.

Demographic differences are modest but systematic. Gender differences are small: respondents coded as gender 2 report slightly higher similarity (M = 3.95) than gender 1 (M = 3.94) and gender 3 (M = 3.91), with largely overlapping distributions in the boxplot. Age effects are clearer: mean similarity increases steadily from Group 1 (13–19, M = 3.92) to Group 3 (28+, M = 3.97). Boxplots show a stable median around 4 across age groups, but older respondents display slightly higher central tendencies.

## Analysis P1: Inferencial stats

first i'm downloading the data set so I don't have to rerun all chuncks of code:

```{r}
saveRDS(Data_RR1, "Data_RR1.rds")
Data_RR1 <- readRDS("Data_RR1.rds")

```

NEW DATA

```{r}
Analysis_RR1 <- Data_RR1 %>%
  filter(!is.na(universe_genre), universe_genre != "Other") %>%
  distinct(resp_id, universe_selected, universe_genre, gender, age_group, enneagram_type, mean_similarity) %>%
  filter(!is.na(mean_similarity))


Analysis_RR1 <-Analysis_RR1 %>%
 mutate(
    universe_genre = factor(universe_genre),
    gender = factor(gender),
    age_group = factor(age_group),
    enneagram_type = factor(enneagram_type)
  )

str(Analysis_RR1)
view(Analysis_RR1)
```

Before linear model testing the first hypothesis, i"ll run first assumption checks, homostecity and normality

Assumption checks (for linear models)

We’ll check assumptions for the core model first. Start with H1 model (genre main effect), then reuse the same checks for interaction models.1A) Fit the baseline linear model (H1)

```{r}
m_h1 <- lm(mean_similarity ~ universe_genre, data = Analysis_RR1)

```

Checking assumptions: Distribution of residuals

```{r}

m_h1$residuals 

hist(m_h1$residuals, breaks = 20,                                    
     main = "Residuals histogram (H1)", xlab = "Residuals")

qqnorm(m_h1$residuals, main = "Normal Q-Q plot (H1)")                 
qqline(m_h1$residuals)                                   

```

The distribution of residuals is approximately symmetric and centered around zero, which is consistent with the assumptions of a General Linear Model.

Checking assumptions: homogeneity of variance

```{r}

plot(m_h1$fitted.values, m_h1$residuals,                                 
     xlab = "Fitted values", ylab = "Residuals",
     main = "Residuals vs Fitted (h1)")
abline(h = 0, lty = 2)              
```

because universe_genre is a binary categorical predictor (Realist vs Fantastical), the model produces only two fitted values (one mean per group). As a result, residuals stack vertically above each fitted value.

The residuals-versus-fitted plot shows two vertical bands corresponding to the two genre categories. This pattern is expected given the categorical predictor. Residuals are centered around zero and exhibit comparable dispersion across fitted values, providing no evidence of heteroscedasticity.

now testing the linear model:

```{r}
summary(m_h1)
```

-   Characters from Realist universes receive similarity scores that are, on average, 0.087 points higher than characters from Fantastical universes.

-   R-squared = 0.0021: Genre explains about 0.2% of the variance in perceived similarity. This is a very small effect size.

Interpretation:

A General Linear Model was estimated to examine the effect of fictional genre on perceived similarity. Results indicate a statistically significant effect of genre, such that characters from realist universes were rated as more similar to respondents than characters from fantastical universes (β = 0.087, p \< .001). However, the magnitude of this effect was small, with genre explaining approximately 0.2% of the variance in perceived similarity (R² = .002). This suggests that while genre is systematically associated with perceived similarity, it accounts for only a limited portion of individual variation in identification with fictional characters.

Visual representation of the model:

```{r}
ggplot(analysis_df, aes(x = universe_genre, y = mean_similarity)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", se = TRUE)
```

References:

Märtens, M., & Reinecke, L. (2021). Personality and preferences for film genres: A large-scale survey study. Personality and Individual Differences, 179, 110927.

Redondo, I., & Holbrook, M. B. (2010). Modeling the appeal of movie features to demographic segments. Journal of Cultural Economics, 34(4), 299–315.

Mares, M.-L., & Sun, Y. (2010). The multiple meanings of age in television viewing. Human Communication Research, 36(3), 372–401.

Mares, M.-L., Cantor, J., & Steinbach, J. B. (2016). Age-related differences in interest in television narratives. Journal of Broadcasting & Electronic Media, 60(1), 1–18.
